Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-07 tex/main.tex
Improvements to explanation of Example 2.

The translation alters $\varphi$ in two main ways. First, it restricts the
quantification on function definitions for $\con{f}$ to the corresponding
uninterpreted type $\fargtype{\con{f}}$ and inserts appropriate selectors
$\fargx{\con{f}}{i}$. Second, it augments $\varphi$ with additional existential
constraints
$\exists a : \fargtype{\con{f}}.\;\, \vecfarg{\con{f}}( a ) \teq \vec t$; these
ensure that the restricted definition for $\con{f}$ covers all relevant tuples
of terms, namely the ones for which an $\con{f}$ application exists and is
relevant to the satisfiability of $\varphi$.
%\jb{what does ``current branch of'' mean? can we drop it / rephrase it / vague
%it up?}

@@

\begin{lemma}\label{lem:conv}
Let $\varphi$ be a formula not containing function definitions,
and $\I$ be a $\extendsig{\Sigma}$-interpretation satisfying $\conv( \varphi )$.
Then, $\I$ satisfies a set of literals $L \cup \absconstraints( L )$ such that $L$ is a set of $\Sigma$-literals that entail $\varphi$.
%There exists a set of ground $\Sigma$-literals $L$ such that $L$ entails $\varphi$, 
%and $\I$ satisfies $L \cup \absconstraints( L )$.
\end{lemma}

@@

The intuition of the above proof is as follows.

For clarity, we demonstrate this in the context of a concrete example.

\begin{example}
Let us revisit the translated formula~(\ref{eq:ex-after}) from
Example~\ref{ex:translation}. This formula is unsatisfiable only if
formula~(\ref{eq:ex-before}) is unsatisfiable since $\fargtype{f}$ can be
interpreted as $\typeint$, and $\farg{f}$ can be interpreted as the identity
function. Conversely, we claim formula~(\ref{eq:ex-after}) is satisfiable
only if formula~(\ref{eq:ex-before}) is satisfiable, noting that the set
$\{ \forallf{\con{s}} x\typ{\typeint}.\;\, \lite\bigl( x \leq 0,\allowbreak\; \con{s}(x)
\teq 0,\; \con{s}( x ) \teq x + \con{s}( x-1 ) \bigr) \}$ is admissible.
%\ct{
%You cannot possibly mean "if and only if" in the two sentences above.
%Once I show that (1) is unsat iff (2) is unsat, I do not need to argue that
%(1) is sat iff (2) is sat. That is immediate.
%}
% You're right, changed both to "only if" --ajr
As stated
in the proof of Theorem~\ref{thm:equi}, the reason is that any model of
formula~(\ref{eq:ex-after}) must satisfy a formula that is closed under
function expansion, hence implying that formula~(\ref{eq:ex-before}) has a
model. In more detail, let $\varphi_\con{s}$ be the formula $\forallf{\con{s}} x.\;\,
\lite\bigl( x \leq 0, \con{s}(x) \teq 0, \con{s}( x ) \teq x + \con{s}( x-1 )
\bigr)$. The rightmost conjunct $\exists a : {\fargtype{\con{s}}}.
\farg{\con{s}}( a ) \teq \con{c}$ ensures that any model of
formula~(\ref{eq:ex-after}) satisfying $\con{s}( \con{c} ) > 100$ also
satisfies $\varphi_{\con{s}}[\con{c}/x]$. The nested existential
$\exists b.\;\, \farg{\con{s}}( b ) \teq \farg{\con{s}}( a )-1$ ensures that
a similar constraint is enforced recursively: when $\varphi_{\con{s}}[t/x]$ is used in part to evaluate $\con{s}( \con{c} )$,
any model of $\varphi_{\con{s}}[t/x]$ must also satisfy $\varphi_{\con{s}}[t-1/x]$ when $t > 0$. Given a model $\I$ of
formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as $14$,
due to the structure of formula~(\ref{eq:ex-after}), it can be shown that $\I$ must satisfy the following formula $\psi$:
%
\begin{equation} \label{eq:fun-closure}
\begin{split}
& \con{s}( \con{c} ) > 100 \wedge %\displaystyle
  \bigwedge\nolimits_{\vthinspace i=0}^{\!13} \bigl( \neg (\con{c}-i \leq 0) \wedge \con{s}( \con{c}-i ) \teq i + \con{s}( \con{c}-i-1 ) \bigr) \\[-\jot]
{\wedge}\; & \con{c}-14 \leq 0 \wedge \con{s}( \con{c}-14 ) \teq 0
\end{split}
\end{equation}
%
This formula is closed under function expansion,
since it contains the $\con{s}$ applications corresponding to
$\con{s}( \con{c}-i )$ for $i = 0, \ldots, 14$. The formula also
entails $\varphi_\con{s}[\con{c}-i/x]$ for $i = 0, \ldots, 14$.
Since $\varphi_\con{s}$ is an admissible definition,
there must exist a $\Sigma$-interpretation satisfying $\psi \wedge \varphi_\con{s}$
which entails formula~(\ref{eq:ex-before}).\xend
\end{example}

>>>>

The translation alters $\varphi$ in two main ways. First, it restricts the
quantification on function definitions for $\con{f}$ to the corresponding
uninterpreted type $\fargtype{\con{f}}$ and inserts appropriate selectors
$\fargx{\con{f}}{i}$. Second, it augments $\varphi$ with additional existential
constraints
$\exists a : \fargtype{\con{f}}.\;\, (\vecfarg{\con{f}}( a ) \teq \vec t)$; these
ensure that the restricted definition for $\con{f}$ covers all relevant tuples
of terms, namely the ones for which an $\con{f}$ application exists and is
relevant to the satisfiability of $\varphi$.
%\jb{what does ``current branch of'' mean? can we drop it / rephrase it / vague
%it up?}

@@

\begin{lemma}\label{lem:conv}
Let $\varphi$ be a formula not containing function definitions,
and $\I$ be a $\extendsig{\Sigma}$-interpretation satisfying $\conv( \varphi )$.
Then, $\I$ satisfies $L \cup \absconstraints( L )$, where $L$ is a set of ground $\Sigma$-literals that entail $\varphi$.
%There exists a set of ground $\Sigma$-literals $L$ such that $L$ entails $\varphi$, 
%and $\I$ satisfies $L \cup \absconstraints( L )$.
\end{lemma}

@@

The intuition of the above proof is as follows.
First, $\conv( \varphi )$ cannot be unsatisfiable when $\varphi$ is satisfiable
since any model of $\varphi$ can be extended in a straightforward way to satisfy $\conv( \varphi )$.
Second, if a model is found for $\conv( \varphi )$,
then the constraints of the form $\exists a : \fargtype{\con{f}}.\;\, (\vecfarg{\con{f}}( a ) \teq \vec t)$
occurring in $\conv( \varphi )$ ensure that this model also satisfies
a $\Sigma$-formula that is closed under function expansion that entails the conjecture of $\varphi$.
Hence when $\DDD$ is admissible, this implies the existence of a model for $\varphi$.
For clarity, we demonstrate the consequences of this theorem in the context of a concrete example.

\begin{example}
Let us revisit the translated formula~(\ref{eq:ex-after}) from
Example~\ref{ex:translation}. 
This formula is unsatisfiable only if
formula~(\ref{eq:ex-before}) is unsatisfiable since $\fargtype{f}$ can be
interpreted as $\typeint$, and $\farg{f}$ can be interpreted as the identity
function. Conversely, we claim formula~(\ref{eq:ex-after}) is satisfiable
only if formula~(\ref{eq:ex-before}) is satisfiable, noting that the set
$\{ \forallf{\con{s}} x\typ{\typeint}.\;\, \varphi_\con{s} \}$ is admissible,
where $\varphi_\con{s}$ is the formula $\lite\bigl( x \leq 0, \con{s}(x) \teq 0, \con{s}( x ) \teq x + \con{s}( x-1 ) \bigr)$.
%\ct{
%You cannot possibly mean "if and only if" in the two sentences above.
%Once I show that (1) is unsat iff (2) is unsat, I do not need to argue that
%(1) is sat iff (2) is sat. That is immediate.
%}
% You're right, changed both to "only if" --ajr
%As stated
%in the proof of Theorem~\ref{thm:equi}, the reason is that any model of
%formula~(\ref{eq:ex-after}) must satisfy a formula that is closed under
%function expansion, hence implying that formula~(\ref{eq:ex-before}) has a
%model.
%The rightmost conjunct $\exists a : {\fargtype{\con{s}}}. (\farg{\con{s}}( a ) \teq \con{c})$ 
%ensures that if $\con{s}( \con{c} ) > 100$ is a constraint that 
%formula~(\ref{eq:ex-after}) satisfying $\con{s}( \con{c} ) > 100$ also
%satisfies $\varphi_{\con{s}}[\con{c}/x]$. 
%The nested existential
%$\exists b.\;\, \farg{\con{s}}( b ) \teq \farg{\con{s}}( a )-1$ ensures that
%a similar constraint is enforced recursively: when $\varphi_{\con{s}}[t/x]$ is used in part to evaluate $\con{s}( \con{c} )$,
%any model of $\varphi_{\con{s}}[t/x]$ must also satisfy $\varphi_{\con{s}}[t-1/x]$ when $t > 0$. 
It is easy to see that any interpretation $\I$ satisfying formula~(\ref{eq:ex-after}) must satisfy $\{ \con{s}( \con{c} ) > 100 \}$ 
(call this set $L_0$),
and moreover must satisfy the set $\absconstraints( L_0 ) = \{ \exists a : {\fargtype{\con{s}}}. (\farg{\con{s}}( a ) \teq \con{c}) \}$.
The latter ensures that $\I$ satisfies a formula that entails both $\con{s}( \con{c} ) > 100$ and $\varphi_{\con{s}}[\con{c}/x]$.
Since $\I$ also satisfies both the translated function definition for $\con{s}$ (e.g. the first conjunct of formula~(\ref{eq:ex-after})) 
and $\absconstraints( L_0 )$, it must also satisfy the formula $\psi_1$:
\[
\lite\bigl( \con{c} \leq 0,\; 
            \con{s}(\con{c}) \teq 0,\;
            \con{s}(\con{c}) \teq \con{c} + \con{s}( \con{c}-1 )
            \wedge \exists b : {\fargtype{\con{s}}}.\;\, (\farg{\con{s}}( b ) \teq \con{c}-1) \bigr) \bigr)
\]

The existential $\exists b : {\fargtype{\con{s}}}.\;\, (\farg{\con{s}}( b ) \teq \con{c}-1)$ ensures that a similar constraint is enforced recursively
when $\I$ satisfies $\neg \con{c} \leq 0$.
Hence, given a model of formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as $14$,
this model must satisfy the following formula $\psi$:
%
\begin{equation} \label{eq:fun-closure}
\begin{split}
& \con{s}( \con{c} ) > 100 \wedge %\displaystyle
  \bigwedge\nolimits_{\vthinspace i=0}^{\!13} \bigl( \neg (\con{c}-i \leq 0) \wedge \con{s}( \con{c}-i ) \teq (\con{c}-i) + \con{s}( (\con{c}-i)-1 ) \bigr) \\[-\jot]
{\wedge}\; & \con{c}-14 \leq 0 \wedge \con{s}( \con{c}-14 ) \teq 0
\end{split}
\end{equation}
%
This formula is closed under function expansion,
since it contains the $\con{s}$ applications corresponding to
$\con{s}( \con{c}-i )$ for $i = 0, \ldots, 14$. The formula also
entails $\varphi_\con{s}[\con{c}-i/x]$ for $i = 0, \ldots, 14$.
Since $\varphi_\con{s}$ is an admissible definition,
there must exist a $\Sigma$-interpretation satisfying $\psi \wedge \varphi_\con{s}$
which entails formula~(\ref{eq:ex-before}).\xend
\end{example}
